The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-19 dictionary.tex
Changes to Applications

\subsection{Smooth commutative group schemes over DVRs}

Let $R$ complete discrete valuation ring $R$ with residue field $\Fq$;
write $\m$ for the maximal ideal of $R$ and $K$ for the quotient 
In this section we use quasicharacter sheaves to geometrize and categorify the group 
of admissible characters of smooth commutative group schemes over $R$.

We will need the Greenberg functor \citelist{\cite{greenberg:61} \cite{greenberg:63a}}:
a functor from the category of schemes locally of finite type
over an Artinian local ring with perfect residue field $k$
to the category of schemes locally of finite type over $k$.
This functor respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components.
See
\citelist{
\cite{demazure-gabriel:GroupesAlgebriques}*{V, \S 4, no. 1}
\cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6}
\cite{nicaise-sebag:motivicSerre}*{\S 2.2}
\cite{bertrapelle-gonzales:Greenberg}*{\S 5}
}
for the definition and fundamental properties of the Greenberg functor as it is used here.
We will write
\[
\Gr_n : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}
\]
for the functor produced by precomposing pullback along $R \to R_n$ with the Greenberg functor.

The surjective ring homomorphism $R_{n+1} \to R_n$ determines a
natural transformation $P_{n} : \Gr_{n+1} \to \Gr_n$ between additive functors.
Set
\begin{align*}
 \GN{X}_n &\ceq \Gr_n(X), \\
 f_n &\ceq P_n(X).
\end{align*}

We list some fundamental properties of $f_n$ and $\GN{X}_n$ for further use:
 \begin{enumerate}
  \labitem{(GN.0)}{truncation}
  $f_n : \GN{X}_{n+1} \to \GN{X}_n$ is a surjective morphism of group schemes over $\Fq$
  \citelist{\cite{bosch-lutkebohmert-reynaud:NeronModels}*{p. 276} \cite{bertrapelle-gonzales:Greenberg}*{Prop. 7.2}}.
  We remark that $f_n : \GN{X}_{n+1} \to \GN{X}_n$ is not induced from a morphism of Néron models.
  \labitem{(GN.1)}{smooth}
  $\GN{T}_n$ is a $\QC$-scheme.
  Note that $\underline{T}$ is a smooth group scheme over $\OK$, by construction.
  Since smoothness is preserved under base change, the group scheme
  $\TT_n \ceq \underline{T} \times_\Spec{\OK} \Spec{\OK/\pK^n}$
  is smooth over an Artinian ring with perfect residue field.
  The Greenberg functor preserves smoothness,
  so $\GN{T}_n$ is a smooth group scheme locally of finite type over $\Fq$.
  The Greenberg functor takes the geometric identity component of $\TT_n$
  to the geometric identity component of $\GN{T}_n$.
  In fact, $\pi_0(\GN{T}_n)$ is independent of $n$, equal to $\comp \ceq \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$.
  Moreover, $\Pi$ is finitely generated \cite{xarles:93a}*{2.18}.
  \labitem{(GN.2)}{identity} $\GN{T}_n^\circ$ is affine of finite type over $\Fq$.
  \labitem{(GN.3)}{points} $\GN{T}_n(\Fq) = \TT(\OK/\pK^n)$, canonically.
\end{enumerate}

\begin{proposition}
The projective limit of the system
\[
\cdots \xrightarrow{f_n} \GN{T}_{n+1} \xrightarrow{f_{n-1}} \cdots \xrightarrow{f_1} \GN{T}_1
\]
exists in the category of group schemes over $\Fq$.
\end{proposition}
\begin{proof}
Since the component group $\comp$ is independent of $n$, the projective system induces a sequence of affine schemes
\begin{equation} \label{eq:Ti}
\cdots \xrightarrow{f^i_n} \GN{T}^i_{n+1} \xrightarrow{f^i_{n-1}} \cdots \xrightarrow{f^i_1} \GN{T}^i_1
\end{equation}
for each $i \in \comp$.
Since the category of affine schemes over $k$ is equivalent to the category of $k$-algebras, which has direct limits \todo{citation},
the limit of \eqref{eq:Ti} exists in the category of affine group schemes over $k$ for each $i \in \comp$.  Write $\GN{T}^i$ for the limit,
and define
\[
\GN{T} \ceq \coprod_{i \in \comp} \GN{T}^i
\]
as a $k$-scheme.  We claim that $\GN{T}$ is a projective limit of the original system, in the category of $k$-schemes.  Suppose that $Z$ is a $k$-scheme
with a system of morphisms $z_n : Z \to \GN{T}_n$.  Let $\alpha : \pi_0(Z) \to \comp$ be the induced map on component groups, which is
independent of $n$.  \todo{the components of $Z$ are not necessarily affine; is the projective limit in the category of affine schemes still the limit in the category of schemes?}

The multiplication maps $m_n : \GN{T}_n \times \GN{T}_n \to \GN{T}_n$ induce maps of affine schemes
$m_n^{i,j} : \GN{T}_n^i \times \GN{T}_n^j \to \GN{T}_n^{i+j},$ inducing maps $m^{i,j} : \GN{T}^i \times \GN{T}$ [under construction]
\end{proof}

Let $X$ be a smooth commutative group scheme over a 
For every $n\in \NN_0$ set $R_n = R/\m^{n+1}$.
We say that a character $\chi: X(R)\to \EEx$ is {\it admissible} if there is some $n \in \NN_0$ such that $\chi$ factors through $X(R) \to X(R_n)$.

$\Hom_\text{ad}(X(R),\EEx)$

\subsection{Smooth commutative group schemes over local fields, with Neron models}

Let $X_K$ be a smooth commutative group scheme over a local field $K$ with residue field $\Fq$. Suppose that $X_K$ admits a Neron model $X$.

>>>>

in this section we use quasicharacter sheaves to geometrize and categorify the group 
of admissible characters of smooth commutative group schemes over local fields that admit Neron models.

Let $R$ complete discrete valuation ring $R$ with residue field $\Fq$.
Write $\m$ for the maximal ideal of $R$ and $K$ for the quotient field for $R$.
Set $R_n = R/\m^{n+1}$ and $S_n = \Spec{R_n}$ and $S = \Spec{R}$; 
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.

\subsection{Greenberg transform}

We will need the Greenberg functor, which is
a functor from the category of schemes 
over an Artinian local ring with perfect residue field
to the category of schemes over the residue field;
See \cite{bertrapelle-gonzales:Greenberg} for the definition and fundamental properties of the Greenberg functor as we use it; other important references include
\citelist{
\cite{greenberg:61}, \cite{greenberg:63a},
\cite{demazure-gabriel:GroupesAlgebriques}*{V, \S 4, no. 1},
\cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6},
\cite{nicaise-sebag:motivicSerre}*{\S 2.2}
}.
This functor respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components.

For any $n\in \NN_0$, we will write
\[
\Gr_n : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}
\]
for the functor produced by composing pullback along $S_n \to S$ with the Greenberg functor. This functor respects \'etale morphisms, smooth morphisms and geometric components, \todo{open immersions? closed immersion?} and
\[
\Gr_n(X)(\Fq) \iso X(R_n),
\]
canonically, for any $X$ lft over $R$.

For any $n\leq m$,  the surjective ring homomorphism $R_{m} \to R_n$ determines a
natural transformation $\Gr_{m} \to \Gr_n$ between additive functors.
Crucially, $(\Gr_{m}\to \Gr_n)(X)$ is an affine morphism of $\Fq$-schemes, for every $R$-scheme $X$ and every $n\leq m$ \cite{bertrapelle-gonzales:Greenberg}*{Prop 4.3}.
This observation is key to the proof that, for any $X$ lft over $S$, the projective limit 
\[
\Gr(X) \ceq \varprojlim_{n\in \NN_0} \Gr_n(X)
\]
taken with respect to the surjective morphisms $(\Gr_{m}\to \Gr_n)(X)$,
exists in the category of group schemes over $\Fq$.
This leads to the definition of what we call the {\it Greenberg transform}:
\[
\Gr : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}.
\]
By construction, the $\Fq$-scheme $\Gr(X)$ comes equipped with morphisms 
\[
P_n(X) : \Gr(X) \to \Gr_n(X),\qquad \forall n\in \NN_0.
\]
The Greenberg transform $\Gr$ respects \'etale morphisms, smooth morphisms  \todo{open immersions? closed immersion?} and geometric components, and
\[
\Gr(X)(\Fq) \iso X(R),
\]
canonically, for any $X$ lft over $R$.

\subsection{Quasicharacter sheaves on the Greenberg transform of a Neron model}

Now suppose $X_K$ is a commutative algebraic group over $K$ 
that admits a Neron model, $X$.
Then $\Gr(X)$ is commutative group scheme over $\Fq$
with component group scheme
$\pi_0(\Gr(X))= \pi_0(X) \times_S S_0$
and
\[
\Gr(X)(\Fq)\iso X_K(K)
\]
canonically.
For every $n\in \NN_0$,
$\Gr_n(X)$ is {\it smooth} commutative group scheme over $\Fq$
with component group scheme
$\pi_0(\Gr_n(X))= \pi_0(X) \times_S S_0$.

A {\it quasicharacter sheaf for $X_K$} is a triple
$\qcs{F}\ceq (\gqcs{F},\mu_\qcs{F},\phi_\qcs{F})$ where
\begin{enumerate}
\item[(0)] $\gqcs{F}$ is a constructible $\ell$-adic sheaf on $\Gr(X)_{\bFq} \ceq \Gr(X)\otimes \bFq$,
\item[(1)]  $\mu_\qcs{F}: \bm^* \gqcs{F} \to \gqcs{F}\boxtimes\gqcs{F}$ is an isomorphism $\ell$-adic sheaves, where $\bm : \Gr(X)_{\bFq} \times \Gr(X)_{\bFq} \to \Gr(X)_{\bFq}$ is multiplication for $\Gr(X)_{\bFq}$,
\item[(2)] $\phi_\qcs{F} : \Frob{\Gr(X)}^* \gqcs{F} \to \gqcs{F}$ is an
  isomorphism of constructible $\ell$-adic sheaves, where $\Frob{\Gr(X)}^* : \Gr(X)_{\bFq}  \to \Gr(X)_{\bFq}$ is Frobenius for $\Gr(X)_{\bFq}$,
\end{enumerate}
such that, for some $n\in \NN_0$,
\begin{enumerate}
\item[(3)]
$\qcs{F} = P_n(X)^*\qcs{L}$ for some quasicharacter sheaf $\qcs{L}$ on $\Gr_n(X)$, where pull-back by $P_n(X)$ is defined as in Proposition~\ref{prop:pullback}.
\end{enumerate}

\begin{theorem}
Let $X_K$ be a commutative algebraic group over $K$ with Neron model $X$ over $R$. 
\begin{enumerate}
\item[(0)] Quasicharacter sheaves for $X_K$ form a rigid monoidal category, $\QC(X_K)$, in a natural way. 
\item[(1)] There is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{X_K} \to \Hom_\text{ad}(X_K(K),\EEx) \to 0
\] 
where $\Hom_\text{ad}(X_K(K),\EEx)$ is the group of characters of $X_K(K)$ that factor through $X(R)\to X(R_n)$, for some $n\in\NN_0$.
\item[(2)] For all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ for $X_K$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
\item[(3)] For all quasicharacter sheaves $\qcs{F}$ for $X_K$,
\[
\Aut(\qcs{L}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
\end{theorem}
